---
title: "The processing function <tt>reh()</tt>"
author: ""
date: ""
output: rmarkdown::html_document
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{The processing function}
  %\VignetteEncoding{UTF-8}
---

This vignette aims to explain inputs, outputs, attributes and methods of the function <tt>remify::reh()</tt>. As example, the `randomREH` data from the same package will be used (documentation available via `?randomREH`).

```{r}
library(remify) # loading library
data(randomREH) # loading data
names(randomREH) # objects inside the list 'randomREH'
```


<i>(Click on a tab pill below to show its content)</i>


# {.tabset .tabset-fade .tabset-pills}


## Input

<b>Input</b> arguments that can be supplied to `reh()` are: `edgelist`, `actors`, `types`, `directed`, `ordinal`, `origin` and `omit_dyad`. 

------

### <b><tt>edgelist</tt></b>

The edgelist must be a `data.frame` or a `matrix` with column names `[time,actor1,actor2,type,weight]` (<i>Note</i>: it is not necessary to supply a column for `type` or `weight` but if they are included in the input edgelist, columns must be named `type` and `weight`, whereas it is compulsory to supply `[time,actor1,actor2]` and use such names for the columns).

```{r}
head(randomREH$edgelist)
```  
------

### <b><tt>actors</tt></b>

It is the vector of actor names (if left unspecified it will be drawn from the input `edgelist`). Their data type can be either `numeric` or `character`. In the `randomREH` data a vector of actor names is provided.

```{r}
randomREH$actors
``` 
------

### <b><tt>types</tt></b>

It is the vector of type names (if left unspecified it will be drawn from the input `edgelist`). The data type can be either `numeric` or `character`. In the `randomREH` data a vector of types is provided.

```{r}
randomREH$types
```
------

### <b><tt>directed</tt></b>

It is a logical `TRUE`/`FALSE` value, indicating whether events are directed (`TRUE`) or not (`FALSE`). If `FALSE`, dyads will be sorted according to their names by following an alphanumerical order (e.g. `[actor1,actor2] = [Colton,Alexander]` will become `[actor1,actor2] = [Alexander,Colton]`).

------

### <b><tt>ordinal</tt></b>
It is a logical `TRUE`/`FALSE` value, indicating whether only the order of events matters in the model. If `TRUE`, then the ordinal likelihood will be used. Otherwise, (when `FALSE`) the standard likelihood for REM will be considered. (<i>C.T. Butts, “A Relational Event Model for Social Action”, Sociological Methodology, 2008</i>) 

------

### <b><tt>origin</tt></b>
If the initial time point $t_0$ is known it can be specified by means of the argument `origin` and it must have the same class of the `time` column in the input `edgelist`. In the case such field is left unspecified (`NULL`), it is set by default to one day/second/time unit earlier than $t_1$ (time of the first observed event). Therefore if `origin = NULL`, when time is measured until seconds then $t_0 = t_1 - 1sec$ , when time is measured until minutes then $t_0 = t_1 - 1min$ and so forth. In the `randomREH` data a $t_0$ is provided.

```{r}
randomREH$origin
``` 

------

### <b><tt>omit_dyad</tt></b>
This argument is useful when certain dyads must be removed from the riskset for specific time windows (e.g. an actor drops out of the network, specific groups of actors cannot interact anymore starting at some time point). Therefore this field makes the riskset suit better to the real data.
`omit_dyad` consists of a list of lists with two objects: a `data.frame` called `dyad`, which specifies the dyads to be removed from the riskset, and `time` which is a vector indicating the time points where to remove the dyads from the riskset.
Consider the example on the `randomREH` data. For instance, we want to modify (shrink) the riskset according to two changes that apply on different time intervals:

1. an event type `conflict` that was no more feasible since $t_r$ (until the last observed time point).

```{r}
randomREH$omit_dyad[[1]]$dyad
```

```{r}
head(randomREH$omit_dyad[[1]]$time) # printing out only a few time points
```

2. two actors `Michaela` and `Zackary` that couldn't interact with anybody else after a specific time point, $t_s$ (until the last observed time point);

```{r}
randomREH$omit_dyad[[2]]$dyad
```

```{r}
head(randomREH$omit_dyad[[2]]$time) # printing out only a few time points
``` 

The object `dyad` will give instructions such that the function will remove from the riskset at the indicated time points all the events where: (1) type is `conflict`, (2) `Michaela` and ``Zackary` are senders or receivers.

The `<NA>` values mean that all the actors/types are considered. Indeed, in the first change where we needed to remove all the events where `conflict` was the type (from $t_r$ to the last time point), we did it by leaving both `actor1` and `actor2` unspecified `<NA>`. 
Therefore, every time one field among (`actor1`,`actor2`,`type`) is left undefined, the omission from the riskset applies to all the values of that field.

### Running the example
```{r}
edgelist_reh <- reh(edgelist = randomREH$edgelist,
                    actors = randomREH$actors,
                    types = randomREH$types, 
                    directed = TRUE, # events are directed
                    ordinal = FALSE, # REM with waiting times
                    origin = randomREH$origin,
                    omit_dyad = randomREH$omit_dyad)
```
## Output

The <b>output</b> of `reh()` is an S3 object of class `reh` and contains the following elements:

```{r}
names(edgelist_reh)
```

------

### <b><tt>M</tt></b>
`M` is the number of observed relational events.

```{r}
edgelist_reh$M
```

------

### <b><tt>N</tt></b>
`N` is the total number of actors that could interact in the network.

```{r}
edgelist_reh$N
```

------

### <b><tt>C</tt></b>
`C` is the number of event types (also referred as the sentiment of the event) that could be observed in the network.

```{r}
edgelist_reh$C
```

------

### <b><tt>D</tt></b>
`D` is the number of possible dyads throughout the network considering also the event types. Therefore:

  * if the network is <i>directed</i>, then $D = N(N-1)C$;
  * if the network is <i>undirected</i>, then $D = (N(N-1)/2)C$.

In other terms, it represents the largest observable size of the riskset.

```{r}
edgelist_reh$D
```

------

### <b><tt>IntereventTime</tt></b>
`intereventTime` is a numeric vector of waiting times between two subsequent events, that is:
$$\begin{bmatrix} t_1 - t_0 \\ t_2 - t_1 \\ \cdots \\ t_M - t_{M-1} \end{bmatrix}$$

```{r}
head(edgelist_reh$intereventTime)
```

------

### <b><tt>edgelist</tt></b>
`edgelist` is a `data.frame` and consists of the converted input edgelist, with columns `[time,actor1,actor2,type,weight]` where `actor1`, `actor2` and `type` are first sorted (alphanumerical order) and then numeric ID's are attributed, ranging in $\left\lbrace 0,\ldots,N-1 \right\rbrace$ for the actors, and in $\left\lbrace 0,\ldots,C-1 \right\rbrace$ for the event types. Where `type` is not defined, its column will consist of a vector of $0$'s. Likewise, if `weight` is not supplied as input, then its corresponding column will be a vector of $1$'s.

```{r}
head(edgelist_reh$edgelist)
```

------

### <b><tt>risksetMatrix</tt></b>
`risksetMatrix` is a matrix of dimensions $[D\times4]$ where columns describe, in order, `[actor1,actor2,type,index]`. ID's are used for actors and types; the `index` column contains values that range in $\left\lbrace 0,\ldots,D-1 \right\rbrace$ where $D$ is the number of possible dyads. This column represents the poisition of the specific dyad `[actor1,actor2,type]` in objects like the array of statistics (where dyads are arranged by column) or in the matrix `rehBinary` (where, by column, user can access the vector of information about time points when the dyad occurred, when it could occur but didn't and when it couldn't occur at all).
`risksetMatrix` is useful in several packages inside `remverse`: for instance, in `remstats` for the computation of statistics.

```{r}
head(edgelist_reh$risksetMatrix)
```


<i>Note:</i> actor ID's and the column `index` are expressed starting from 0 (not 1), in turn, following the indexing convention in C++. Therefore, if they have to be used as position indices in R programming language, they must be increased by 1.

------

### <b><tt>risksetCube</tt></b>
`risksetCube` is a 3-dimensional array of size $[N \times N \times C]$: the first dimension (rows) refers to `actor1`, the second (columns) refers to `actor2` and the third (slice) refers to `type`. Therefore, a specific combination of `[actor1,actor2,type]` in terms of their ID's returns an index which is the same as the one in the fourth column of `risksetMatrix`.

```{r}
edgelist_reh$risksetCube[1:10,1:10,1] # this code prints out only the first slice (event type with ID = 0) for just the first 10 actors with ID's ranging in [0,1,...,9]
```


<i>Note:</i> actor ID's and indices are expressed starting from 0 (not 1), in turn, following the indexing convention in C++. Therefore, if they have to be used as position indices in R programming language, they must be increased by 1.

------

### <b><tt>rehBinary</tt></b>
`rehBinary` is a `matrix` of size $[M \times D]$ and by row (time point) it consists of a vector where each element assumes value:
   * $1$ if the specific `[actor1,actor2,type]` occurred at that time point;
   * $0$ if  the specific `[actor1,actor2,type]` did't occur but could have occurred (thus it is present in the riskset at that time point);
   * $-1$ if  the specific `[actor1,actor2,type]` didn't occur and couldn't have occurred (thus it is removed from the riskset at that time point).
   
```{r}
edgelist_reh$rehBinary[1:10,1:10] # this code prints out only the first ten events (by row) and the first 10 combinations of [actor1,actor2,type] (by column). Same combinations of risksetMatrix[1:10,]
```

The columns correspond to a specific combination combination of `[actor1,actor2,type]`, and if the user wants to know with dyad and type that column refers to, she has to use the `risksetCube` with actors ID's and type ID in order to find the right column index (ID's must be increased by 1 if working in R).

## Attributes


The <b>attributes</b> of an `reh` object are:

```{r}
names(attributes(edgelist_reh))
```

------

### <b><tt>names</tt></b>
`names` is the vector of the names of the output objects which are discussed already in section 'Output'.

```{r}
attr(edgelist_reh, "names")
```

------

### <b><tt>class</tt></b>
The `class` attribute returns the class name of the object, that is `reh`.

```{r}
attr(edgelist_reh, "class")
```

------

### <b><tt>with_type</tt></b>
`with_type` is a logical `TRUE`/`FALSE` value indicating whether more than one event type is observed in the network (`TRUE`) or not (`FALSE`).

```{r}
attr(edgelist_reh, "with_type")
```

------

### <b><tt>weighted</tt></b>
`weighted` is a logical `TRUE`/`FALSE` value indicating whether relational events have weights (`TRUE`) or not (`FALSE`).

```{r}
attr(edgelist_reh, "weighted")
```

------

### <b><tt>directed</tt></b>
`directed` is a logical `TRUE`/`FALSE` value indicating whether we know (`TRUE`) for each event whom originated the action (sender) and whom was the target (receiver) of it, or we don't know (`FALSE`) the source and the target of an event but only the actors that were involved in it.

```{r}
attr(edgelist_reh, "directed")
```

------

### <b><tt>ordinal</tt></b>
`ordinal` is a logical `TRUE`/`FALSE` value indicating whether in the model we want to consider the waiting times between events (`FALSE`) or not and consider only the order of relational events (`TRUE`).

```{r}
attr(edgelist_reh, "ordinal")
```

------

### <b><tt>riskset</tt></b>
`riskset` describes the type of riskset that resulted from the processing of the data. If `omit_dyad` is provided, this means that the riskset is going to change for certain time windows and then `riskset = "dynamic"`. Otherwise, when `omit_dyad = NULL`, the riskset is assumed to be the largest observable one (of dimensions $D$) and it maintains the same shape throughout the network (`riskset = "static"`).  

```{r}
attr(edgelist_reh, "riskset")
```

------

### <b><tt>dictionary</tt></b>
`dictionary` is a list of two `data.frame`'s: `actors` and `types`.

  - `actors` has two columns: the first with actor names (`actorName`) sorted according to their alphanumerical order, the second with their corresponding ID's (`actorID`), ranging in $\left\lbrace 0,\ldots,N-1 \right\rbrace$;
  - `types` has two columns: the first with type names (`typeName`) sorted according to their alphanumerical order, the second with their corresponding ID's (`typeID`), ranging in $\left\lbrace 0,\ldots,C-1 \right\rbrace$; 

```{r}
attr(edgelist_reh, "dictionary")
```

------

### <b><tt>time</tt></b>
`time` is a list of three objects: `class`, `value` and `origin`.

  - `class` returns the class of the column `time` provided in the input `edgelist`;
  - `value` is a `data.frame` of size $[M \times 2]$ where the first column is the `time` variable supplied in the input `edgelist` and the second column is the `intereventTime` (already discussed in section 'Output');
  - `origin` is the input argument already discussed in section 'Input'.

```{r}
str(attr(edgelist_reh, "time")) # printing out only the str() of the attribute since the data.frame `value` is large
```



## Methods

The available <b>methods</b> for an `reh` object are: `summary`, `dim`, `print`, `View`, `head`, `tail`, `getRiskset`, `actorName`, `typeName`, `actorID`, `typeID`, `dyad.info`.

------

### <b><tt>summary</tt></b>
`summary(reh)` prints out a brief summary of the relational network data.

```{r}
summary(edgelist_reh)
```

------

### <b><tt>dim</tt></b>
`dim(reh)` returns some useful dimensions characterizing the network, such as: number of events, number of actors, number of event types, number of dyads. 

```{r}
dim(edgelist_reh)
```

------

### <b><tt>print</tt></b>
`print(reh)` prints out the whole output `edgelist` on the console.
```{r}
#print(edgelist_reh)
```

------

### <b><tt>View</tt></b>
`View(reh)` opens a window with a view of the whole output `edgelist`.

```{r}
#View(edgelist_reh)
```

------

### <b><tt>head</tt></b>
`head(reh)` returns the first six rows of the output `edgelist`.

```{r}
head(edgelist_reh)
```

------

### <b><tt>tail</tt></b>
`tail(reh)` returns the last six rows of the output `edgelist`.

```{r}
tail(edgelist_reh)
```

------

### <b><tt>getRiskset</tt></b>
`getRiskset(reh)` returns a logical matrix of size $[M \times D]$ where by row (time point), dyads that could occur at a specific time point assume value `TRUE`, `FALSE` otherwise. Below, only a small portion of the matrix is shown. 

```{r}
getRiskset(edgelist_reh)[1:10,1:10] # printing out only the first 10 rows (time points) and the first 10 columns (dyads)
```

------

### <b><tt>actorName</tt></b>
`actorName(reh,actorID)` by supplying one or more `actorID`'s it returns the corresponding (input) names.

```{r}
actorName(reh = edgelist_reh, actorID = c(0,12,19))
```

------

### <b><tt>typeName</tt></b>
`typeName(reh,typeID)` by supplying one or more `typeID`'s it returns the corresponding (input) names.

```{r}
typeName(reh = edgelist_reh, typeID = c(0,2))
```

------

### <b><tt>actorID</tt></b>
`actorID(reh,actorName)` by supplying one or more `actorName`'s it returns the corresponding ID's.

```{r}
actorID(reh = edgelist_reh, actorName = c("Michaela","Alexander","Lexy"))
```

------

### <b><tt>typeID</tt></b>
`typeID(reh,typeName)` by supplying one or more `typeName`'s it returns the corresponding ID's.

```{r}
typeID(reh = edgelist_reh, typeName = "cooperation")
```

------

### <b><tt>dyad.count</tt></b>
`dyad.count(reh,actor1,actor2,type,begin,end)` is a method that returns the count of a dyad in the network. User can supply `[actor1,actor2,type]` in such way that if one of the three argument is left undefined, the count will be made across all the values for the undefined fields. Furthermore, the user can supply two time indices, `begin` and `end`, thus the count will be performed only according to a specific time window.


#### Example 1
Counts of events where `Michaela` is the `actor1` and `Richard` is `actor2`. In the example the network is directed, thus we can refer to `actor1` as <i>sender</i> and to `actor2` as <i>receiver</i>.

```{r}
# whole network
dyad.count(reh = edgelist_reh, actor1 = "Michaela", actor2 = "Richard") 

# whole network and event type 'conflict'
dyad.count(reh = edgelist_reh, actor1 = "Michaela", actor2 = "Richard", type = "conflict") 

# first time window (from the 1st time point to the 3999th)
dyad.count(reh = edgelist_reh, actor1 = "Michaela", actor2 = "Richard", begin = 1, end = 3999) 

# second time window (from the 4000th time point to the last one)
dyad.count(reh = edgelist_reh, actor1 = "Michaela", actor2 = "Richard", begin = 4000) 
```

#### Example 2
Counts of events with a specific `actor1` or `actor2`

```{r}
# counts of dyads where Michaela is the sender
dyad.count(reh = edgelist_reh, actor1 = "Michaela") 

# counts of dyads where Richard is the receiver
dyad.count(reh = edgelist_reh, actor2 = "Richard")
```

#### Example 3
Counts of events with a specific event type.
```{r}
# counts for type 'conflict' (only first five actors)
dyad.count(reh = edgelist_reh, type = "conflict")[1:5,1:5]

# counts for type 'cooperation' (only first five actors)
dyad.count(reh = edgelist_reh, type = "cooperation")[1:5,1:5]
```

