---
title: "Transform processed relational event sequences to different formats"
author: ""
package: remify
date: ""
output: 
  rmarkdown::html_vignette:
    css: "remify-theme.css"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Transform processed relational event sequences to different formats}
  %\VignetteEncoding{UTF-8}
---

---

<i> This vignette explains input arguments, output structure and usage of the function `remify::rehshape()`. </i>

---

### Aim

`remify::rehshape() ` transforms a `remify` object into another object with a structure that is suitable to external packages. The function can return the data inputs required by the functions:

-  `relevent::rem()` 
-  `relevent::rem.dyad()` 

Both functions are available inside the [relevent](https://CRAN.R-project.org/package=relevent) package (Butts C., 2023).

---

### Input

The input arguments of `remify::rehshape()` are:

 - `data`, the processed relational event history (S3 object of class `remify`)
 - `output_format`, a character value that indicates to which output format the input `data` should be converted. This argument can assume two values: `"relevent-rem"` , `"relevent-rem.dyad"` (default is `"relevent-rem"`)
 - `ncores`, number of threads used to parallelize internal routines (default is `1L`)
 - `optional_arguments`, vector of arguments names from `relevent::rem` or `relevent::rem.dyad()` that the user might want to process and have in the output object of rehshape (e.g., the pre-computed structures required by `relevent::rem.dyad()`, such as `acl`, `cumideg`, etc.) - _this feature will be available in a future version of remify_ -

---

### Output
The output structure of the function is different according to the chosen `output_format`:

- If `output_format = "relevent-rem"`, then the output is an S3 object of class `relevent-rem`, which contains: 
  - `eventlist`, a matrix of two columns: observed dyads in the first column, vector of time in the second column
  - `supplist`, a logical matrix of dimensions [rows = number of events, columns = number of dyads]. The matrix indicates at each time point (by row) whether each dyad was at risk (`TRUE`), or not (`FALSE`)
  - `timing`, is a character that can assume two values: `"interval"` (which uses the inter-event time in the model), or `"ordinal"` (which only considers the event order in the model)

- If `output_format = "relevent-rem.dyad"`, then the output is an S3 object of class `relevent-rem.dyad`, which contains: 
  - `edgelist`, a matrix of three columns: the time (or order) of the events in the first column, the sender and the receiver of the relational event, respectively, in the second and third column
  - `n`, is the number of actors in the relational event network (senders and receivers)
  - `ordinal`, is a logical (`TRUE`/`FALSE`) value which indicates whether the likelihood should be 'ordinal' (`TRUE`) or 'interval' (`FALSE`)

---

### Usage
To explain the usage of the function `remify::rehshape()`, we consider the example edgelist available with the data `randomREH`. First, we process the edgelist with `remify::remify()`.
```{r}
library(remify)
data(randomREH) 
reh_remify <- remify::remify(edgelist = randomREH$edgelist, model = "tie")
reh_remify
```

Then, we can transform the `remify` object to any of the possible output formats:

-  from `remify` to `relevent-rem`:
```{r}
reh_rem <- remify::rehshape(data = reh_remify, 
                      output_format = c("relevent-rem"))
names(reh_rem)                     
```

-  from `remify` to `relevent-rem.dyad`:
```{r}
reh_rem.dyad <- remify::rehshape(data = reh_remify, 
                      output_format = c("relevent-rem.dyad"))
names(reh_rem.dyad)                     
```
---
