---
title: "Process a Relational Event History"
author: ""
package: remify
date: ""
output: 
  rmarkdown::html_document:
    theme: spacelab
    highlight: pygments
    code_folding: show
    css: "remify-theme.css"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Process a Relational Event History}
  %\VignetteEncoding{UTF-8}
---

---

<i>This vignette explains aim, input, output, attributes and methods of the function <tt>`remify::remify()`</tt>.</i>

---


# {.tabset .tabset-fade .tabset-pills}


## Aim
The objective of <tt>remify::remify()</tt> is to process raw relational event sequences supplied by the user along with other inputs that characterize the data (actors' names, event types' names, starting time point of the event sequence, set of interactions to be excluded from the risk set at specific time points, etc.).
The internal routines will process the structure of the input event sequence into a new one, providing also objects that will be used by the packages in 'remverse'.

As example, we will use the data `randomREH` (documentation available via `?randomREH`).

```{r}
library(remify) # loading library
data(randomREH) # loading data
names(randomREH) # objects inside the list 'randomREH'
```


---


## Input

<b>Input</b> arguments that can be supplied to `remify()` are: `edgelist`,`directed`, `ordinal`,`model`, `actors`, `types`, `riskset`, `origin`, `omit_dyad` and `ncores`. 

------

### <b><tt>edgelist</tt></b>

The edgelist must be a `data.frame` with three mandatory columns: the time of the interaction in the first column, and the two actors forming the dyad in the second and third column. The naming of the first three columns is not required but the order must be `[time,actor1,actor2]`. For directed networks, the second column of the edgelist will be the column of the sender and the third column will be the one of the receiver. For undirected network, the order of second and third column is ignored. Optional columns that can be supplied are `type` and `weight` and they have to be named accordingly. However, type and weight must be defined after the third column.

```{r}
head(randomREH$edgelist)
```  
------

### <b><tt>directed</tt></b>

It is a logical `TRUE`/`FALSE` value, indicating whether events are directed (`TRUE`) or not (`FALSE`). If `FALSE`, dyads will be sorted according to their names by following an alphanumeric order (e.g. `[actor1,actor2] = ["Colton","Alexander"]` will become `[actor1,actor2] = ["Alexander","Colton"]`).

------

### <b><tt>ordinal</tt></b>
It is a logical `TRUE`/`FALSE` value, indicating whether only the order of events matters in the model (when `TRUE`) or also the waiting times between event must be taken into account (when `FALSE`).
Based on the value of this argument, the processing of the time variable is carried out differently and the `remstimate` package will use either the _ordinal_ (if `ordinal` is `TRUE`) or the _interval_ (if `ordinal` is `FALSE`) time likelihood. 

------


### <b><tt>model</tt></b>
Whether the model of interest is tie-oriented or actor-oriented, the argument `model` can be specified either as `"tie"` or as `"actor"`. This argument affects the structure of the output as to the risk set: in the case of an actor-oriented model with changing risk set, the processed risk set will consist of two objects, senders' and dyads' risk sets, in a tie-oriented modeling, the output risk set will consist of only the riskset on the dyads. If no dynamic risk set is defined (`omit_dyad = NULL`), the argument `model` won't affect the processing of the event sequence.

------

### <b><tt>actors</tt></b>

It is the vector of actor names (if left unspecified, names will be taken from the input `edgelist`). Their data type can be either `numeric` or `character`. In the `randomREH` data, a vector of actor names is provided.

```{r}
randomREH$actors
``` 
------

This argument is useful especially when the user wants to include actors that could interact during the study but didn't actually take part in any interaction. Therefore, by not including their names via the argument `actors` will exclude them from the risk set.

### <b><tt>types</tt></b>

It is the vector of type names (if left unspecified, names will be taken from the input `edgelist`). The data type can be either `numeric` or `character`. In the `randomREH` data a vector of types is provided.

```{r}
randomREH$types
```

------

### <b><tt>riskset</tt></b>
The `riskset` argument specifies the type of risk set that should be used for the input data. It can assume three different values: `"full"`, `"active"` and `"manual"` (with `"full"` being the default value). The risk set is `"full"`, if all the possible dyadic interactions given the number of actors in the network should be considered at risk throughout the event sequence. The risk set is `"active"`, if it is only made of only the observed dyadic interactions and it assumed that they are the only dyads at risk for the entire event sequence. Finally, the risk set is `"manual"`, if it has a time-varying structure, that is, when one or more actors should be excluded from the network until a specific time point because the join the network later, or vice versa, when one or more actors should be excluded from the risk set from a time point and on because they leave the network before the end of the sequence. In other words, a manual risk set is defined whenever one or more dyadic interactions cannot occur during one or more timespans. If the `riskset` argument is `"manual"`, then the user should supply a list object with the risk set modifications to the `omit_dyad` argument.
More details about risk set definitions are provided in `vignette(topic = "riskset", package = "remify")`.

------

### <b><tt>origin</tt></b>
The initial time ($t_0$) of an event sequence is the time at which the network of actors starts being observed and any event occurring from that time point and on is annotated. This information is not always known and it depends on the study design. If the $t_0$ of an event sequence is known, it can be specified by the argument `origin` and it must have the same class of the `time` column in the input `edgelist`. If the argument is left unspecified (`NULL`), it will be set by default to one time unit earlier than $t_1$ (time of the first observed event). For instance, when the waiting time is measured in seconds then $t_0 = t_1 - 1sec$ , when the waiting time is measured in days then $t_0 = t_1 - 1day$ and so forth. In the `randomREH` data a $t_0$ is provided.

```{r}
randomREH$origin
``` 

------

### <b><tt>omit_dyad</tt></b>
This argument is required when `riskset="manual"`, therefore one or more dyads must be omitted from the risk set for specific time windows (e.g. an actor drops out of the network, specific groups of actors cannot interact anymore starting from some time point). 
The `omit_dyad` input is a list of lists. Each list refers to one risk set modification and must have two objects: a `data.frame` called `dyad`, where dyads to be remove are specified by row in the format `actor1, actor2, type`, and `time` which is a vector of two values defining the first and last time point of the time window in which such dyads couldn't occur.
Consider the example on the `randomREH` data. For instance, we want to modify (shrink) the risk set according to two changes that apply on different time intervals:

1. an event type `conflict` that cannot be observed since a specific time point until the end of the observation period.

```{r}
randomREH$omit_dyad[[1]]$time # start and stop time point defining the time window of interest
```
```{r}
randomREH$omit_dyad[[1]]$dyad # dyads to be removed from the time points defined by the interval in `time`
```

2. two actors `Michaela` and `Zackary` that cannot interact with anybody else after a specific time point until the last observed time point.


```{r}
randomREH$omit_dyad[[2]]$time # start and stop time point defining the time window of interest
```
```{r}
randomREH$omit_dyad[[2]]$dyad # dyads to be removed from the time points defined by the interval in `time`
```

The object `dyad` will give instructions such that the function will remove from the risk set at the indicated time windows all the events where: (1) type is `conflict`, (2) `Michaela` and ``Zackary` are senders or receivers of a relational event.

The `<NA>` values mean that all the actors/types are considered in that field. Indeed, in the first change where we needed to remove all the events where `conflict` was the type, we did it by leaving both `actor1` and `actor2` unspecified `<NA>`. 
Therefore, every time one field among (`actor1`,`actor2`,`type`) is left undefined, the omission from the risk set applies to all the possible values of that field.

Furthermore, the internal routines in `remify::remify()`, given the dyads specified via the `omit_dyad` input, always operate a _shrinkage_ of the risk set starting from the "full" risk set (all the possible dyads given the number of actors and information about directed/undirected events). Therefore, the dyads described in the `omit_dyad` list must be based on the observed set of actors (actors from the event sequence) and on the input vector of actors (supplied via the argument `actors`). If the user wants to _add_ a set of actors to the risk set for a specific timespan, this can be done by: first making sure that the names of such actors are found either in the input `edgelist` or specified in the input `actors`, then the user can define an `omit_dyad` object that specifies the time window(s) in which that set of actors is not at risk, so that they are at risk only during a desired time window.


------

### Running the example
```{r}
edgelist_reh <- remify(edgelist = randomREH$edgelist,
                    directed = TRUE, # events are directed
                    ordinal = FALSE, # model with waiting times
                    model = "tie", # tie-oriented modeling
                    actors = randomREH$actors,
                    types = randomREH$types, 
                    riskset = "manual",
                    origin = randomREH$origin,
                    omit_dyad = randomREH$omit_dyad)
                    
```

## Output

The <b>output</b> of `remify()` is an S3 object of class `remify` and contains the following elements:

```{r}
names(edgelist_reh)
```

------

### <b><tt>M</tt></b>
`M` is the number of observed time points. If there are events occurring at the same time point, then `M` will count the number of unique time points and the number of events will be returned by `E` (see below). If all events occurred at different time points, then `M` will correspond also to the number of events and `E` will not be defined.

```{r}
edgelist_reh$M
```

------

### <b><tt>E</tt></b>
`E` is the number of observed events. `remify()` returns the number of observed events only if there are events that occurred at the same time point. If all observed events occurred at different time points, then `remify()` will not return `E`.

------

### <b><tt>N</tt></b>
`N` is the total number of actors that could interact in the network.

```{r}
edgelist_reh$N
```

------

### <b><tt>C</tt></b>
`C` is the number of event types (also referred as the sentiment of the event) that could be observed in the network. If no event types are present in the network or only one event type is defined, then the output object `C` will be `NULL`.

```{r}
edgelist_reh$C
```

------

### <b><tt>D</tt></b> and <b><tt>activeD</tt></b>
`D` is the number of possible dyads considering the number of actors and also the number of event types (if measured). It represents the largest size of the risk set (full risk set size). The number of dyads `D` can change based on whether relational events are directed or undirected:
    * if the network is <i>directed</i>, then $D = N*(N-1)*C$;
    * if the network is <i>undirected</i>, then $D = (N*(N-1)/2)*C$.

If the input argument `riskset` is `"active"`, then in the output object there will also be the number `activeD`, which is the size of the active risk set (active dyads observed in the event history).

```{r}
edgelist_reh$D
```

------

### <b><tt>intereventTime</tt></b>
`intereventTime` is a numeric vector of waiting times between two subsequent events, that is:
$$\begin{bmatrix} t_1 - t_0 \\ t_2 - t_1 \\ \cdots \\ t_M - t_{M-1} \end{bmatrix}$$

```{r}
head(edgelist_reh$intereventTime)
```

The vector of waiting times is available only if the sequence is processed for interval likelihood (`ordinal = FALSE`). In the case of ordinal likelihood, `intereventTime` is NULL.

------

### <b><tt>edgelist</tt></b>
`edgelist` is a `data.frame` and consists in the original input edgelist, with columns `[time,actor1,actor2,type,weight]` where events by-row are re-ordered if the time variable in the input edgelist was not correctly sorted. If `type` or `weight` are not supplied as input, then the output edgelist will not incude their columns.

```{r}
head(edgelist_reh$edgelist)
```

------

### <b><tt>omit_dyad</tt></b>

When `riskset = "manual"`, `omit_dyad` must be supplied to the processing function, then the output object will contain the processed list under the same name. In the case of tie-oriented modeling, the list consists of two objects: a vector named `time` and a matrix named `riskset`.

- `riskset` is a 1/0's matrix where all the possible risk set modifications are described by row, and the columns identify the dyads ($D$ columns). The number of rows depends on the number of risk set modifications occurring in the event sequence, thus it remains variable (and it is not necessarly the same as the number of risk set modifications declared with the input `omit_dyad`, because they can also occur on no/partially/totally overlapping time windows that are processed internally).

- `time` is a vector of row indices that for each time point indicates which modification of the risk set (row index in the matrix `riskset`) is observed.

```{r}
edgelist_reh$omit_dyad$riskset[,1:10] # printing out the risk set modifications of only the first 10 columns (dyads). A total number of 2 modifications of the risk set are observed (by row)
```

The processed object `omit_dyad` (inside any `remify` object) will be required by other packages in 'remverse'. Given that such packages have function written in C++, the row indices in the vector start at 0 (indicating row 1 in the matrix above) and they assume value -1 when no risk set alteration is observed.

```{r}
edgelist_reh$omit_dyad$time[1:10] # printing out the first 10 time points. We can see that in none of the 10 time points any modification takes place (-1)
```

In the case of actor-oriented modeling with changing risk set, the `omit_dyad` output list consists of three objects: a vector named `time` (same vector as explained above) and two risk set matrices (`senderRiskset` and `riskset`), one for the sender and one for the dyads. Both risk set matrices follow the same structure as the matrix `riskset` in the tie-oriented modeling.
Whenever `omit_dyad` is not supplied to `remify()` then its output in the `remify` object will be empty.


## Attributes

Consider now an `"active"` risk set and process the same edgelist:
```{r}
edgelist_reh <- remify(edgelist = randomREH$edgelist,
                    directed = TRUE, # events are directed
                    ordinal = FALSE, # model with waiting times
                    model = "tie", # tie-oriented modeling
                    actors = randomREH$actors,
                    types = randomREH$types, 
                    riskset = "active",
                    origin = randomREH$origin)
                    
```


The <b>attributes</b> of a `remify` object are:



```{r}
names(attributes(edgelist_reh))
```

------

### <b><tt>names</tt></b>
`names` is the vector of the names of the output objects which are discussed already in section 'Output'.

```{r}
attr(edgelist_reh, "names")
```

------

### <b><tt>class</tt></b>
The `class` attribute returns the class name of the object, that is `remify`.

```{r}
attr(edgelist_reh, "class")
```

------

### <b><tt>with_type</tt></b>
`with_type` is a logical `TRUE`/`FALSE` value indicating whether more than one event type is observed in the network (`TRUE`) or not (`FALSE`).

```{r}
attr(edgelist_reh, "with_type")
```

------

### <b><tt>weighted</tt></b>
`weighted` is a logical `TRUE`/`FALSE` value indicating whether relational events have weights (`TRUE`) or not (`FALSE`).

```{r}
attr(edgelist_reh, "weighted")
```

------

### <b><tt>directed</tt></b>
`directed` is a logical `TRUE`/`FALSE` value indicating whether we know (`TRUE`) for each event whom originated the action (sender) and whom was the target (receiver) of it, or we don't know (`FALSE`) the source and the target of an event but only the actors that were involved in it.

```{r}
attr(edgelist_reh, "directed")
```

------

### <b><tt>ordinal</tt></b>
`ordinal` is a logical `TRUE`/`FALSE` value indicating whether in the model we want to consider the waiting times between events (`FALSE`) or not and consider only the order of the relational events (`TRUE`).

```{r}
attr(edgelist_reh, "ordinal")
```

------

### <b><tt>model</tt></b>
`model` describes whether the output of `remify()` is suitable for the actor-oriented model (`model = "actor"`) or for the tie-oriented model (`model = "tie"`).

```{r}
attr(edgelist_reh, "model")
```

------

### <b><tt>riskset</tt></b>
`riskset` returns the type of risk set that was chosen in the processing of the data. Possible values are: `"full"`, `"manual"` and `"active"`.

```{r}
attr(edgelist_reh, "riskset")
```

------

### <b><tt>dictionary</tt></b>
`dictionary` is a list of two `data.frame`'s: `actors` and `types`.

  - `actors` has two columns: the first with actor names (`actorName`) sorted according to their alphanumerical order, the second with their corresponding ID's (`actorID`), ranging from $1$ to $N$ (with $N$ being the number of actors);
  - `types` has two columns: the first with type names (`typeName`) sorted according to their alphanumerical order, the second with their corresponding ID's (`typeID`), ranging from $1$ to $C$ (with $C$ being the number of event types);

```{r}
attr(edgelist_reh, "dictionary")
```

For more details about how the ID is assigned to actors and types, see `utils::vignette(topic = "riskset", package = "remify")`

------

### <b><tt>origin</tt></b>
`origin` is the value of the input argument with the same name.

```{r}
str(attr(edgelist_reh, "origin")) # printing out only the str() of the attribute since the data.frame `value` is large
```

------

### <b><tt>ncores</tt></b>
`ncores` is the number of threads used in the parallelization of internal routines in `remify::remify()`.

```{r}
attr(edgelist_reh, "ncores")
```

------

### <b><tt>dyadID</tt></b>
`dyadID` is a list of vectors containing the IDs of the observed dyads per time point. The IDs range between 1 and `D` (`edgelist_reh$D`). For more details about how the ID is assigned to a specific triple of `[actor1,actor2,type]`, see `vignette(topic = "riskset", package = "remify")`.

```{r}
attr(edgelist_reh, "dyadID")[[1]] # printing out dyads ID's observed at the first time point
```

------


### <b><tt>actor1ID</tt></b>
`actor1ID` is a list of vectors containing the IDs of the observed actor1's/senders per time point. The IDs range between 1 and `N` (`edgelist_reh$N`). 

```{r}
attr(edgelist_reh, "actor1ID")[[1]] # printing out the actor1's/senders ID's observed at the first time point
```

------


### <b><tt>actor2ID</tt></b>
`actor2ID` is a list of vectors containing the IDs of the observed actor2's/receivers per time point. The IDs range between 1 and `N` (`edgelist_reh$N`). 

```{r}
attr(edgelist_reh, "actor2ID")[[1]] # printing out the actor2's/receivers ID's observed at the first time point
```

------

### <b><tt>typeID</tt></b>
`typeID` is a list of vectors containing the IDs of the observed types per time point. The IDs range between 1 and `C` (`edgelist_reh$C`). 

```{r}
attr(edgelist_reh, "typeID")[[1]] # printing out the types ID's observed at the first time point
```

------

### <b><tt>dyadIDactive</tt></b>
`dyadIDactive` is a list of vectors containing the IDs of the active set of dyads. This attribute is available only when `riskset = "active"`.

```{r}
attr(edgelist_reh, "dyadIDactive")[[1]] # printing out the ID's of the active dyads at the first time point
```


------

### <b><tt>evenly_spaced_interevent_time</tt> and <tt>indices_simultaneous_events</tt></b>
`evenly_spaced_interevent_time` is the vector of processed waiting times when in the relational event sequence two or more events are observed at the same time point. This attribute will be `NULL` when there are no simultaneous events in the sequence. The evenly spaced interevent time assumes that if two or more events occurred at the same time, the waiting time since the previous (different) time point is evenly spread across the simultaneous events. For instance, if we observe five events at time `10` and the previous (different) time point is `4`, their raw waiting time will be calculated as below.

```{r}
time_points <- c(4,10,10,10,10,10)
waiting_times <- diff(time_points) # waiting_times: [1] 6 0 0 0 0 calculated as t[m]-t[m-1]
```
The waiting time of the second event from the first event is `6` (calculated as the difference of the two time points, `10-4=6`). The value of the waiting time is `0` for the other events that are simultaneous (i.e. `10-10=0`). In this scenario, `remify` spreads the waiting time `6` for the five simultaneous events as,

```{r}
rep(waiting_times[1]/5,5)# 5 is the number of events in the example observed at the same time (10)
```

`indices_simultaneous_events` is a vector of indices from the input `edgelist` indicating which rows of the edgelist correspond to simultaneous events (excluding the first simultaneous event). This attribute will be `NULL` when there are no simultaneous events in the sequence. Considering the example above, if there are five simultaneous events at time `10`, `remify` will return the following set of indices
```{r}
time_points <- c(4,10,10,10,10,10)
diff(time_points) # waiting times calculated as t[m]-t[m-1]
which(diff(time_points)==0) # indices of simultaneous events, excluding the first simultaneous event 
```
`remify`, doesn't remove the first simultaneous event for each case of two or more simultaneous events in the sequence because that time point must be processed to group the simultaneous events as occurred at a specific time point.
```{r}
# attr(edgelist_reh, "evenly_spaced_interevent_time") 
# attr(edgelist_reh, "indices_simultaneous_events")
```


## Methods

The available <b>methods</b> for a `remify` object are: `print`, `summary`, `dim`, `getRiskset`, `getActorName`, `getTypeName`, `getDyad`, `getActorID`, `getTypeID`, `getDyadID` and `plot`.

------

### <b><tt>print() and summary()</tt></b>
Both `print()` and `summary()` print out a brief summary of the relational network data.

```{r}
summary(edgelist_reh) # same output as `print(edgelist_reh)` or just `edgelist_reh`
```

------

### <b><tt>dim()</tt></b>
`dim()` returns some useful dimensions characterizing the network, such as: number of events, number of actors, number of event types (omitted in the case of networks with one or no event types), largest number of possible dyads ($D$) and, finally, number of active dyads (`activeD`, showed only if `riskset = "active"`). 

```{r}
dim(edgelist_reh)
```

------

### <b><tt>getRiskset()</tt></b>

If risk set is `"active"` or `"manual"`, then `getRiskset()` returns the processed risk set matrix (or the two dynamic risk sets if actor-oriented modeling) that are explained in the 'Output' section.

```{r}
getRiskset(x = edgelist_reh)$riskset[,1:10] # printing out the risk set modifications of only the first 10 columns (dyads). A total number of 2 modifications of the risk set are observed (by row)
```
------


### <b><tt>getActorName()</tt></b>
`getActorName(x,actorID)`: by supplying one or more `actorID`'s (between `1` and `edgelist_reh$N`) it returns the corresponding (input) names.

```{r}
getActorName(x = edgelist_reh, actorID = c(1,13,20))
```

------

### <b><tt>getTypeName()</tt></b>
`getTypeName(x,typeID)`: by supplying one or more `typeID`'s (between `1` and `edgelist_reh$C`) it returns the corresponding (input) names.

```{r}
getTypeName(x = edgelist_reh, typeID = c(1,3))
```

------

### <b><tt>getDyad()</tt></b>
`getDyad(x, dyadID, active)`: by supplying one or more `dyadID`'s it returns a `data.frame` with the dyad composition as "actor1", "actor2" and "type" (if type is present). If `active = FALSE`, then the method expects that the vector `dyadID` consist of ID's of dyads from the full risk set (between `1` and `edgelist_reh$D`). If the risk set in the processed remify object is `"active"`, then it is also possible to get the dyad composition from the ID of the active dyads by declaring the argument `active = TRUE` and supplying the vector of ID's of the active dyads to `dyadID` (ranging between `1` and `edgelist_reh$activeD`).

```{r}
getDyad(x = edgelist_reh, dyadID = c(1,10,100), active = FALSE)
```


------

### <b><tt>getActorID()</tt></b>
`getActorID(x,actorName)`: by supplying one or more `actorName`'s it returns the corresponding ID's.

```{r}
getActorID(x = edgelist_reh, actorName = c("Michaela","Alexander","Lexy"))
```

------

### <b><tt>getTypeID()</tt></b>
`getTypeID(x,typeName)`: by supplying one or more `typeName`'s (if types are present in the event sequence) it returns the corresponding ID's.

```{r}
getTypeID(x = edgelist_reh, typeName = "cooperation")
```

------

### <b><tt>getDyadID()</tt></b>
`getDyadID(x,actor1,actor2,tpye)`: by supplying a vector of names as to actor1, actor2 and type it returns the corresponding dyad ID. The names to supply are the original input names of the edgelist before the processing.
```{r}
getDyadID(x = edgelist_reh, actor1 = "Alexander", actor2 = "Charles", type = "cooperation")
```

The method can work with only one input dyad of the form [`actor1`,`actor2`,`type`]. The argument `type` can be left `NULL` if the sequence has no event types (attribute `"with_type"` is `FALSE`).

------

### <b><tt>plot()</tt></b>
`plot()` returns different descriptive plots:

1. histogram of the inter-event times
2. activity plot, which is a tile plot where the color of the tiles is proportional to the count of the directed (or undirected) dyad. In-degree and out-degree activity line plots are plotted on the sides (or the total-degree on the top side, if the network is undirected).
3. for directed networks two plots of normalized out-degree and in-degree values per actor (ranging in $[0,1]$) over a set of `n_intervals` (evenly spaced). For undirected networks, only one plot of normalized total-degree over the `n_intervals` (also here values ranging in [0,1]). The normalization is calculated in each interval as the $$\frac{(degree-min(degree))}{(max(degree)-min(degree))}$$ for each actor considering minimum and maximum degree (in-, out- or total-, degree) observed in the interval (opacity and size of the points is proportional to the normalized measure) 
4. four plots: (i) number of events (# events) per time interval, (ii) proportion of observed dyads (# dyads / `x$D`) per time interval, (iii) and (iv) (for directed network only) proportion of active senders and receivers per time interval (calculated as # senders/ `x$N` and # receiver/`x$N` per interval) 
5. two networks: (i) network of events where edges are considered undirected (edges' opacity is proportional to the counts of the undirected events, vertices' opacity is proportional to the total-degree of the actors), (ii) visualization of directed network (edges' opacity is proportional to the counts of the directed events, vertices' opacity is proportional to the in-degree of the actors).


```{r, out.width="50%", fig.align = "center", dev=c("jpeg")}
op <- par(no.readonly = TRUE)
par(mai=rep(0.8,4), cex.main=0.9, cex.axis=0.75)
plot(x=edgelist_reh,which=1,n_intervals=13) # histogram of inter-event times
plot(x=edgelist_reh,which=2,n_intervals=13) # tile plot (counts of dyadic events) with in-/out- degree of actors on the sides
plot(x=edgelist_reh,which=3,n_intervals=13) # (normalized) in-degree and out-degree of actors
plot(x=edgelist_reh,which=4,n_intervals=13) # per time interval: number of events, proportion of observed dyads, proportion of active senders and active receivers
plot(x=edgelist_reh,which=5,n_intervals=13,igraph.edge.color="#cfcece",igraph.vertex.color="#7bbfef") # networks
par(op)
```

The plots above slightly vary in the case of undirected networks:

```{r}
edgelist_reh <- remify(edgelist = randomREH$edgelist,
                    directed = FALSE, # events are now considered undirected
                    model = "tie")    
#op <- par(no.readonly = TRUE)
#par(mai, rep(0.8,4), cex.main=0.9, cex.axis=0.75)           
#plot(x=edgelist_reh,which=1:5,n_intervals=13)
#par(op) 
```